---
title: "Code Enforcement"
output:
  pdf_document:
    pandoc_args: [
      "--latex-engine=xelatex"
    ]
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(lubridate)
library(jsonlite)
library(knitr)
library(ggplot2)
library(leaflet)
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
options (scipen=4)
cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999", "#000000", "#8856a7")

# QUESTION
# appears that many addresses don't have eventIDs
# does the property database contain all properties
# regardless of if they have violations?
# QUESTION
# PropertyID vs ParID
# Are they consistently coded between both
# Do some tables only use PropertyID
# which should we use for joins?

#### Things we want to analyze
#
# number of inspections per month
# number of resolved and unresolved per month
#   # new ones plus outstanding
# days of compliance
#   # this month
#   # this year
#   # how long particular violation at partiulcar address
# muliple violations by address
# maybe multiple violations by owner
#   # possibly look into linking this across municipalites
# open violations
#   # date resolved
#   # violation type
#   # address
# maybe map (might put on website instead)

## Filter dataframes for municipality
## Could probably figure out how to do this
## by creating a function and then using `apply` later
muni.data <- filter(muni.data, muniname == municipality)
closed <- filter(closed, muniname == municipality)
monthCases <- filter(monthCases, muniname == municipality)
monthClosed <- filter(monthClosed, muniname == municipality)
monthOpen <- filter(monthOpen, muniname == municipality)
muni.year <- filter(muni.year, muniname == municipality)
open <- filter(open, muniname == municipality)
```

# *`r municipality` `r monthName` `r yearChar` Report*

### This Month's Activity

This month, Enforcement Officers

 - Opened `r length(unique(monthCases$EventID))` new cases
 - `r length(unique(monthOpen$EventID))` of which are still open and
 - `r length(unique(monthClosed$EventID))` are closed, with the remaining cases handled informally.
 - `r sum(muni.data$ComplianceDate > reportMonthDate & muni.data$ComplianceDate < nextMonth, na.rm=T)` cases were marked as resolved in the past month.

```{r month, echo=FALSE, warning=FALSE, message=FALSE}
month.count <- monthCases %>%
  group_by(Violation) %>%
  summarise(
    Count = n()
  )%>%
  filter(!is.na(Violation), Violation!="") %>%
  arrange(desc(Count))
```

```{r monthBar, message=FALSE, warning=FALSE, fig.height=5.5}
# reorder factor levels correctly
month.count$Violation <- factor(month.count$Violation, 
                                levels = month.count$Violation[order(month.count$Count)])

head(month.count, 8) %>%
  ggplot(aes(Violation, Count)) + 
  geom_bar(aes(fill = Violation), stat = "identity") + 
  scale_fill_manual(values=cbPalette) + 
  ggtitle("Top Violations This Month") + 
  xlab("") +
  theme(legend.justification=c(1,1), legend.position=c(0.65,1), axis.text.x=element_blank())
```

### Past Twelve Months

```{r yearSetup, echo=FALSE, warning=FALSE, message=FALSE}
year.count <- muni.year %>%
  filter(min(EventDate) > (reportMonthDate - months(12))) %>%
  # group by Violation to get the violation counts
  group_by(Violation) %>%
  summarise(
    Count = n(),
    Open = sum(OverallStatus=="Open"),
    Closed = sum(OverallStatus=="Closed")
  )%>%
  filter(!is.na(Violation), Violation!="") %>%
  arrange(desc(Count))
```

This year, Enforcement Officers handled

 - `r length(unique(muni.year$EventID))` unique cases
 - `r table(muni.year$OverallStatus)[2]` of which are still open,
 - `r table(muni.year$OverallStatus)[1]` are closed, with the remaining cases handled informally.

```{r year, warning=FALSE, message=FALSE}
# plot count of violations by month
muni.year %>%
  mutate(Month = month(EventDate)) %>%
  group_by(Month) %>%
  summarise(
    Count = n(),
    Open = sum(OverallStatus=="Open"),
    Closed = sum(OverallStatus=="Closed")
  ) %>%
  ggplot(aes(Month, Count)) +
  geom_bar(stat = "identity", fill="#a63603") +
  labs(y = "Violations Count", title="Count by Month") + 
  # label months with names instead of numbers
  scale_x_continuous(breaks = c(1:12), labels = c("Jan", "Feb", "Mar", "Apr",
                                                  "May","Jun", "Jul", "Aug", 
                                                  "Sep", "Oct","Nov", "Dec"))
```

## Top Violations This Year
```{r yearBar, message=FALSE, warning=FALSE, fig.height=6}
# reorder factor levels correctly
year.count$Violation <- factor(year.count$Violation, 
                                levels = year.count$Violation[order(year.count$Count)])

head(year.count, 8) %>%
  ggplot(aes(Violation, Count)) + 
  geom_bar(aes(fill = Violation), stat = "identity") + 
  scale_fill_manual(values=cbPalette) + 
  ggtitle("Top Violations Twelve Months") + 
  xlab("") +
  theme(legend.position="bottom", legend.direction="vertical", axis.text.x=element_blank())
```

## All Violations This Year
```{r yearTable, warning=FALSE, message=FALSE}
kable(year.count)
```

This graphic shows the top violations, and how they fluctuate over the course of the year.


```{r yearByType, message=FALSE, warning=FALSE}
# sort violation types by order and then excerpt top violations
violation.names <- sort(table(muni.year$Violation), decreasing = TRUE)
top.violations <- head(violation.names, 6)

muni.year %>%
  filter(Violation %in% names(top.violations)) %>%
  mutate(Month = month(EventDate)) %>%
  group_by(Month, Violation) %>%
  summarise(
    Count = n()
  ) %>%
  ggplot(aes(x=Month, y=Count, group=Violation, colour=Violation)) +
  geom_line(size = 1, alpha=0.8) +
  expand_limits(x = 0, y = 0) +
  labs(y = "Violations Count",
       title="Overall Top Violations Types by Month") +
  scale_colour_manual(values=cbPalette) +
  #theme(legend.justification=c(1,1), legend.position=c(1,1)) +
  theme(legend.position="bottom", legend.direction="vertical") +
  # label months with names instead of numbers
  scale_x_continuous(breaks = c(1:12), labels = c("Jan", "Feb", "Mar", "Apr",
                                                  "May","Jun", "Jul", "Aug", 
                                                  "Sep", "Oct","Nov", "Dec"))
```

### This Year's Properties With 10+ Violations

This table lists the PropertyIDs for those with more than 10 violations, and date the violation was resolved, where available.

```{r year_address, message=FALSE, warning=FALSE}
# only last month's violations in same by address format
# want to filter to only those that are still open but can't get
# that filter to work above
# so instead trying filtering by just 10+ violations
# also could filter by (Violation %in% names(top.violations)) but
# that appears to be getting the same list
year.byParcelID <- muni.year %>%
  group_by(PropertyID) %>%
  summarise(
    Count = n(),
    FirstDate = min(EventDate),
    DateResolved = max(DateResolved)
    #Status = OverallStatus[max(EventDate)]
    #TimeOpen = ComplianceDate - FirstDate
  ) %>%
  arrange(desc(Count)) %>%
  filter(Count > 10)

kable(year.byParcelID)
```


```{r map}
# map.data <- filter(muni.data, !is.na(Violation), !is.na(lat))
# leaflet() %>%
#       # use the CartoDB map for more subdued map tile colors
#       addProviderTiles("CartoDB.Positron") %>%
#       setView(lng = -79.8, lat = 40.4212, zoom = 12) %>%
#       addCircleMarkers(data = map.data,
#                        lng = ~long, lat = ~lat, 
#                        popup = paste("Violation:", map.data$Violation,
#                                      "<br> Initial Date:", map.data$EventDate),
#                  # the marker clusters gives us a nice visual 
#                  # effect where when we zoom out points cluster
#                  # together, and as we zoom in they are pulled apart
#                  # showing the individual markers.
#                  #clusterOptions = markerClusterOptions()
#                  radius = 5,
#                  stroke = TRUE,
#                  color = "#323232",
#                  weight = 3,
#                  fillColor = "#7fcdbb",
#                  fillOpacity = 0.7
#                  ) 
```
